---
title: "Cookbook"
author: "Kirthana Sukumaran"
date: "September 05, 2019"
output:
html_document:
df_print: paged
---
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
graphics.off()
#install.packages("readxl")
library(readxl)
#devtools::install_github("herveabdi/PTCA4CATA") # you only need this line of code if you haven't downloaded the package
#install.packages("gridExtra")
#install.packages("ggplotify")
#install.packages("grid")
library(devtools)
library(ExPosition)
library(InPosition)
library(corrplot)
library(ggplot2)
library(PTCA4CATA)
#install.packages("dplyr")
#install.packages("tidyverse")
library(dplyr)
# The following three functions allow you to arrange your plots in Rmarkdown
library(gridExtra)
library(ggplotify)
library(grid)
library(tidyverse)
```

# 1. Principal Component Analysis

**Data table:** PCA is used to analyze one table of quantitative data. 

**Goal:** PCA computes new variables called principal components which are obtained as linear combinations of the original variables to find new variables that maximizes the variance od the data. 


**Key ideas**
1. Principle components are orthogonal to each other.

**Interpretation**
   
    1. Factor scores are the coordinates of the row observations. They are interpreted by the
       distances between them, and their distance from the origin. 

    2. Loadings describe the column variables. Loadings are interpreted by the angle between
       them, and their distance from the origin.

    3. The distance from the origin is important in both maps, because squared distance from
       the mean is inertia (variance, information; see sum of squares as in ANOVA/regression).
       Because of the Pythagorean Theorem, the total information contributed by a data point
       (its squared distance to the origin) is also equal to the sum of its squared factor
       scores.


##Dataset: Survey of Autobiographical Mememory

The data was collected by Baycrest Institute at University of Toronto. Participants with different memory scores took several questionnaire. 

Participants were asked to rate the extent to which a particular item applied to their memory in general, using a **5-point Likert scale**  (1- completely disagree, 2-4 - intermediate degrees of agreement/disagreement, 5 completely agree).   

There are 153 obseravtions(**rows**) which represents the participants who answer to 26(**Columns**) questions that comprised of 8 Episodic memory based questions 6 Semantic memory questions, 6 Spatial memory based questions and 6 Prospective memory related questions. 

The subjects include both men and women with their ages in the range of 18-84 years which are also mentioned as age and sex variable. A survey based measure of AM is also used to caatogorize the participants into two groups- High memory, Normal Memory.

```{r include=FALSE}
# load and read dataset....
SAMfulldataset <- read_excel("C:\\Users\\KIRTHANA\\Desktop\\UT dallas\\ABDI LAB\\SAM_OSIQ\\SAM_only.xlsx")
head(SAMfulldataset)

d <-SAMfulldataset 
d_active <- d[!(d$Active == FALSE),]

```
Dataset Cleaning

Steps:
1. Check for NA or incomplete data in the dataset and remove them if they exist.
2. Remove Mysterious Memory Groups : participants with conflicting responses
3. Flitering data : Use only Numeric data to perform PCA

Preprocessing the Data

1. Centering: Refers to subtracting mean of each column from each of its points. 
2. Scaling: Normalization (Since SAM data consists of likert scale they all range from 0-5. There exixts homogenity in units of the data. SO refrain from scaling     the data.


```{r altering_data, include=FALSE}
# Removing mysterious memory groups

#d_filter<-filter(d, Active==FALSE)
d_use<-d[!(d$Active == FALSE),]

# Converting dataset to only numerical data 
d_Num<- d_use[ , c(6:31)]
Mean_col<- colMeans(d_Num)
# center the data without normalizing it
d_centered<- scale(d_Num, center = TRUE, scale = FALSE)



```


## Looking at the data pattern

**Correlation Plot**

**What it does?**
The corrplot package is a graphical display of a correlation matrix, confidence interval. It also contains some algorithms to do matrix reordering. In addition, corrplot is good at details, including choosing color, text labels, color labels, layout, etc.

**Analyzing the plot**
There are seven visualization methods (parameter method) in corrplot package, named "circle", "square", "ellipse", "number", "shade", "color", "pie".

Positive correlations are displayed in blue and negative correlations in red color. Color intensity and the size of the circle are proportional to the correlation coefficients.



```{r, echo = TRUE}

E <- c("#5f9ea0","#5f9ea0","#5f9ea0","#5f9ea0","#5f9ea0","#5f9ea0","#5f9ea0","#5f9ea0")
S <- c("#ff7f24", "#ff7f24", "#ff7f24", "#ff7f24", "#ff7f24", "#ff7f24")
P <- c("#76ee00","#76ee00","#76ee00","#76ee00","#76ee00","#76ee00")
Fu<- c("#305ABF","#305ABF","#305ABF","#305ABF","#305ABF","#305ABF")

color.vector <- as.matrix(c(E,Fu,P,S)) 

#colors for groups by rows

col4row <- d_active$memoryGroups
col4row <- recode(col4row, 
                  Low = 'orange',
                  Norm = 'tomato', 
                  High = 'darkred'
                     )

cor.res <- cor(d_centered)
corrplot.pca <-corrplot(cor.res, method = "color", tl.cex = .5, tl.col = color.vector,
         addCoef.col = "black", number.digits = 0, number.cex = .4, 
         cl.pos = 'b', cl.cex = .3, 
         addCoefasPercent = TRUE,
         col = colorRampPalette(c("darkred", "white","midnightblue"))(20)) 

corrplot.pca<-recordPlot()
``` 

## Performing epPCA
* `center = TRUE`: substracts the mean from each column 
* `scale = FALSE`: after centering (or not), scales each column to have a sum of squares of 1 (see the help for different scaling options)
* `DESIGN`: colors the observations (rows)
* `graphs = FALSE`: this gives you plots from `epPCA`, but make sure to flag it `FALSE` for Rmarkdown to run correctly 

Note: We run the epPCA and epPCA.inference package by passing the data containing only quantitative variables and also the design variables(colors for plot)

```{r, echo=TRUE}

# Using numerical data and disregarding age and gender factor
d_PCA<- d_use[ , c(6:31)]

res_pca<- epPCA(
          d_PCA, 
          center = TRUE, 
          scale = FALSE, 
          DESIGN = d_use$memoryGroups, # Memory Groups  of participants based on AM
          graphs = FALSE
          )
```

### Scree Plot

**What it does**
The scree plot shows the eigenvalues, the amount of information on each component. The number of components (the dimensionality of the factor space) is min(nrow(DATA), ncol(DATA)) minus 1. 
**Analysing the Plot**
Here, 8 columns give 7 components. The scree plot is used to determine how many of the components should be interpreted. 

```{r scree plot}
SAM.scree <- PlotScree(ev = res_pca$ExPosition.Data$eigs)
SAM.scree <- recordPlot() 
# you need this line to be able to save them to ppt.
```
## PCA inference

The inference battery package includes permutation and bootstrap tests.

```{r, echo=TRUE, fig.height=3, fig.width=3, message=FALSE}
res_pcaInf <- epPCA.inference.battery(d_PCA, center = TRUE,
                                      scale = FALSE,
                                      DESIGN = d_use$memoryGroups, 
                                      graphs = FALSE)


```
### Scree Plot
Show the results from permutation with Scree plot (i.e., color the significant components) by adding the estimated p-values to the `PlotScree` function.
```{r, echo=TRUE}

my.scree <- PlotScree(ev = res_pcaInf$Fixed.Data$ExPosition.Data$eigs,
                      p.ev = res_pcaInf$Inference.Data$components$p.vals)

my.scree <- recordPlot()

```
### Testing the eigenvalues
```{r}

zeDim = 1
pH1 <- prettyHist(
           distribution = res_pcaInf$Inference.Data$components$eigs.perm[,zeDim], 
           observed = res_pcaInf$Fixed.Data$ExPosition.Data$eigs[zeDim], 
           xlim = c(0, 3000), # needs to be set by hand
           breaks = 10,
           border = "white", 
           main = paste0("Permutation Test for Eigenvalue ",zeDim),
           xlab = paste0("Eigenvalue ",zeDim), 
           ylab = "", 
           counts = FALSE, 
           cutoffs = c( 0.975))

pH1

pH1 <- recordPlot()
  
  
zeDim = 2
pH2 <- prettyHist(
           distribution = res_pcaInf$Inference.Data$components$eigs.perm[,zeDim], 
           observed = res_pcaInf$Fixed.Data$ExPosition.Data$eigs[zeDim], 
           xlim = c(0, 1800), # needs to be set by hand
           breaks = 10,
           border = "white", 
           main = paste0("Permutation Test for Eigenvalue ",zeDim),
           xlab = paste0("Eigenvalue ",zeDim), 
           ylab = "", 
           counts = FALSE, 
           cutoffs = c(0.975))


pH2




zeDim = 3
pH3 <- prettyHist(
           distribution = res_pcaInf$Inference.Data$components$eigs.perm[,zeDim], 
           observed = res_pcaInf$Fixed.Data$ExPosition.Data$eigs[zeDim], 
           xlim = c(0, 1000), # needs to be set by hand
           
           breaks = 10,
           border = "white", 
           main = paste0("Permutation Test for Eigenvalue ",zeDim),
           xlab = paste0("Eigenvalue ",zeDim), 
           ylab = "", 
           counts = FALSE, 
           cutoffs = c(0.975))

pH3






```


### Factor scores

*Row Factor scores*
F = XQ = P*delta
Projections of the observations onto the principal components.

```{r factor_scores}
my.fi.plot <- createFactorMap(res_pcaInf$Fixed.Data$ExPosition.Data$fi, # data
                            title = "SAM Row Factor Scores", # title of the plot
                            axis1 = 2, axis2 = 3, # which component for x and y axes                      
                            display.labels = FALSE,
                            pch = 19, # the shape of the dots (google `pch`)
                            cex = 2, # the size of the dots
                            text.cex = 2.5, # the size of the text
                            alpha.points = 0.3,
                            col.points = col4row, # color of the dots
                            col.labels = col4row # color for labels of dots
                            )

fi.labels <- createxyLabels.gen(2,3,
                            lambda = res_pcaInf$Fixed.Data$ExPosition.Data$eigs,
                            tau = round(res_pcaInf$Fixed.Data$ExPosition.Data$t),
                            axisName = "Component "
                            )
fi.plot <- my.fi.plot$zeMap + fi.labels # you need this line to be able to save them in the end
fi.plot
```

Color for each group:
```{r}
# get index for the first row of each group
grp.ind <- order(d_use$memoryGroups)[!duplicated(sort(d_use$memoryGroups))]
grp.col <- res_pcaInf$Fixed.Data$Plotting.Data$fi.col[grp.ind] # get the color
grp.name <- d_use$memoryGroups[grp.ind] # get the corresponding groups
names(grp.col) <- grp.name
```


#### With group means
```{r, echo=TRUE}

group.mean <- PTCA4CATA::getMeans(res_pcaInf$Fixed.Data$ExPosition.Data$fi,
                        d_use$memoryGroups)

col4Means <- recode(rownames(group.mean), 
                    Low = 'orange',
                    High = 'darkred', 
                    Norm = 'tomato2', 
                  )
names(col4Means) <- rownames(group.mean)


```


```{r fimeanplot}

fi.mean.plot <- createFactorMap(group.mean,
                                alpha.points = 1,
                                axis1 = 2,
                                axis2 = 3,
                                display.labels = TRUE,
                                col.points = col4Means,
                                col.labels = col4Means,
                                pch = 17,
                                cex = 3,
                                text.cex = 3
                                )

fi.WithMean <- my.fi.plot$zeMap_background + 
               my.fi.plot$zeMap_dots +
               fi.mean.plot$zeMap_dots +
               fi.mean.plot$zeMap_text + 
               fi.labels
fi.WithMean
```

#### Tolerance interval
The spread of the factor scores based on their corresponding groups
```{r}
TIplot <- MakeToleranceIntervals(res_pcaInf$Fixed.Data$ExPosition.Data$fi,
                            design = as.factor(d_use$memoryGroups),
                            # line below is needed
                            names.of.factors =  c("Dimension 2","Dimension 3"), # needed 
                            col = col4Means,
                            line.size = .50, 
                            line.type = 3,
                            alpha.ellipse = .2,
                            alpha.line    = .4,
                            p.level       = .95)

fi.WithMeanTI <- my.fi.plot$zeMap_background +
                 my.fi.plot$zeMap_dots+
                 fi.mean.plot$zeMap_dots + 
                 fi.mean.plot$zeMap_text+ 
                 TIplot + fi.labels

fi.WithMeanTI
```


#### Bootstrap interval
We can also add the bootstrap interval for the group means to see if these group means are significantly different.

```{r echo=TRUE}
# Depend on the size of your data, this might take a while
fi.boot <- Boot4Mean(res_pcaInf$Fixed.Data$ExPosition.Data$fi[,1:3],
                     design = d_use$memoryGroups,
                     niter = 1000)

for (i in 1:ncol(fi.boot$BootCube)){
  colnames(fi.boot$BootCube)[i] <- paste("Dimension", i)
}

```

```{r echo=TRUE}

axis1=2
axis2=3

bootCI4mean <- MakeCIEllipses(fi.boot$BootCube[,c(2:3),], # get the first two components
  names.of.factors = paste0("Dimension ", c(2,3)),
                              col = col4Means, )

fi.WithMeanCI <- my.fi.plot$zeMap_background + 
                 fi.labels + bootCI4mean+
                 my.fi.plot$zeMap_dots +                  
                 fi.mean.plot$zeMap_dots + 
                 fi.mean.plot$zeMap_text 
                 
fi.WithMeanCI

```

### Loadings

Loadings describe the similarity (angular distance) between the variables. Loadings show how the input variables relate to each other. Loadings also show which variables are important for a certain component.

```{r eval=FALSE, include=FALSE}
# using the means of column to plot the Loadings

res_pca<- epPCA(d_PCA, 
          center = TRUE, 
          scale = FALSE, 
          DESIGN = d_use$memoryGroups, 
          graphs = FALSE 
          )

E1<- recode(res_pca$Plotting.Data$fj.col[c(1:8),1] ,   "mediumorchid4" =	"#5f9ea0")


S1<-recode (res_pca$Plotting.Data$fj.col[c(9:14),1] , "mediumorchid4" =	"#ff7f24")


P1<-recode(res_pca$Plotting.Data$fj.col[c(15:20),1], "mediumorchid4" = "#76ee00")

F1<-recode(res_pca$Plotting.Data$fj.col[(21:26)], "mediumorchid4" ="#305ABF")

                                         
res_pca$Plotting.Data$fj.col<-c(E1,S1,P1,F1) 

cor.loading <- (cor(d_PCA, res_pca$ExPosition.Data$fi))
colnames(cor.loading) <- rownames(cor.loading)

```

```{r eval=FALSE, include=FALSE}
loading.plot <- createFactorMap(cor.loading,
                                constraints = list(minx = -1, miny = -1,
                                                   maxx = 1, maxy = 1))
  
LoadingMapWithCircles <- loading.plot$zeMap + 
                         addArrows(cor.loading,color= res_pca$Plotting.Data$fj.col)+
                         addCircleOfCor() + 
                         xlab("Component 1") + 
                         ylab("Component 2")                     
LoadingMapWithCircles
```


```{r echo=TRUE}

res_pcaInf$Plotting.Data$fj.col<-color.vector 

cor.loading <- cor(d_PCA, res_pcaInf$Fixed.Data$ExPosition.Data$fi)
colnames(cor.loading)<- rownames(cor.loading)

loading.plot <- createFactorMap(cor.loading,
                                constraints = list(minx = -1, miny = -1, maxx = 1, maxy = 1),col.labels = color.vector,
                              
                                col.points = color.vector )

LoadingMapWithCircles <- loading.plot$zeMap + 
                         addArrows(cor.loading, color = color.vector) + 
                         addCircleOfCor() + 
                         xlab("Component 1") +
                         ylab("Component 2")


LoadingMapWithCircles
```
*Column Factor Scores*
You can also include the variance of each component and plot the factor scores for the columns (i.e., the variables). The following looks at the column factor scores for dimension 1cand 2.

```{r echo=TRUE}
res_pcaInf$Plotting.Data$fj.col<-color.vector 

my.fj.plot <- createFactorMap(res_pcaInf$Fixed.Data$ExPosition.Data$fj, # data
                            title = "SAM Column Factor Scores", # title of the plot
                            axis1 =1 , axis2 = 2, # which component for x and y axes
                            pch = 19, # the shape of the dots (google `pch`)
                            cex = 3, # the size of the dots
                            text.cex = 3, # the size of the text
                            col.points = color.vector , # color of the dots
                            col.labels = color.vector , # color for labels of dots
                            )

fj.plot <- my.fj.plot$zeMap + fi.labels 
fj.plot
```


Similarly looking at column factor scores for dimension 2 and 3

```{r echo=FALSE}

my.fj.plot2 <- createFactorMap(res_pcaInf$Fixed.Data$ExPosition.Data$fj, # data
                            title = "SAM Column Factor Scores", # title of the plot
                            axis1 = 2, axis2 = 3, # which component for x and y axes
                            pch = 19, # the shape of the dots (google `pch`)
                            cex = 3, # the size of the dots
                            text.cex = 3, # the size of the text
                            col.points = color.vector , # color of the dots
                            col.labels = color.vector , # color for labels of dots
                            )

fi23.labels <- createxyLabels.gen(2,3,
                            lambda = res_pcaInf$Fixed.Data$ExPosition.Data$eigs,
                            tau = round(res_pcaInf$Fixed.Data$ExPosition.Data$t),
                            axisName = "Component "
                            )
fj.plot2 <- my.fj.plot2$zeMap + fi23.labels # you need this line to be able to save them in the end
fj.plot2

```

* Component 1: Episodic memory

* Component 2: Semantic memory V/S Spatial memory and Future memory

* Component 3: Future memory V/s Spatial


### Contributions of columns and their Bootstrap Ratios

* Contribution Bar plots : How the variables contribute to each component by plotting
                           the contributions with a line that represents the threshold                            we compared them to.

* Bootstrap ratio Bar plots: Checks the significance of contribution of variables to
                             the priciple components.


```{r include=FALSE}
res_pcaInf$Fixed.Data$Plotting.Data$fj.col <- color.vector
signed.ctrJ <- res_pcaInf$Fixed.Data$ExPosition.Data$cj * sign(res_pcaInf$Fixed.Data$ExPosition.Data$fj)


# plot contributions for component 1
ctrJ.1 <- PrettyBarPlot2(signed.ctrJ[,1],
                         threshold = 1 / NROW(signed.ctrJ),
                         font.size = 4,
                         color4bar = gplots::col2hex(res_pcaInf$Fixed.Data$Plotting.Data$fj.col), # we need hex code
                         ylab = 'Contributions',
                         ylim = c(1.2*min(signed.ctrJ), 1.2*max(signed.ctrJ))
) + ggtitle("Contribution barplots", subtitle = 'Component 1: Variable Contributions (Signed)')

# plot contributions for component 2
ctrJ.2 <- PrettyBarPlot2(signed.ctrJ[,2],
                         threshold = 1 / NROW(signed.ctrJ),
                         font.size = 4,
                         color4bar = gplots::col2hex(res_pcaInf$Fixed.Data$Plotting.Data$fj.col), # we need hex code
                         ylab = 'Contributions',
                         ylim = c(1.2*min(signed.ctrJ), 1.2*max(signed.ctrJ))
) + ggtitle("",subtitle = 'Component 2: Variable Contributions (Signed)')


BR <- res_pcaInf$Inference.Data$fj.boots$tests$boot.ratios
laDim = 1

# Plot the bootstrap ratios for Dimension 1
ba001.BR1 <- PrettyBarPlot2(BR[,laDim],
                        threshold = 2,
                        font.size = 4,
                   color4bar = gplots::col2hex(res_pcaInf$Fixed.Data$Plotting.Data$fj.col), # we need hex code
                  ylab = 'Bootstrap ratios',
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle("Bootstrap ratios", subtitle = paste0('Component ', laDim))


# Plot the bootstrap ratios for Dimension 2
laDim = 2
ba002.BR2 <- PrettyBarPlot2(BR[,laDim],
                        threshold = 2,
                        font.size = 4,
                   color4bar = gplots::col2hex(res_pcaInf$Fixed.Data$Plotting.Data$fj.col), # we need hex code
                  ylab = 'Bootstrap ratios'
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle("",subtitle = paste0('Component ', laDim))

```



```{r, include=FALSE}
# plot contributions for component 3

ctrJ.3 <- PrettyBarPlot2(signed.ctrJ[,3],
                         threshold = 1 / NROW(signed.ctrJ),
                         font.size = 4,
                         color4bar = gplots::col2hex(res_pcaInf$Fixed.Data$Plotting.Data$fj.col), # we need hex code
                         ylab = 'Contributions',
                         ylim = c(1.2*min(signed.ctrJ), 1.2*max(signed.ctrJ))
) + ggtitle("",subtitle = 'Component 3: Variable Contributions (Signed)')





# Plot the bootstrap ratios for Dimension 3

laDim = 3
ba003.BR3 <- PrettyBarPlot2(BR[,laDim],
                        threshold = 2,
                        font.size = 4,
                   color4bar = gplots::col2hex(res_pcaInf$Fixed.Data$Plotting.Data$fj.col), # we need hex code
                  ylab = 'Bootstrap ratios'
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle("",subtitle = paste0('Component ', laDim))





```

```{r  echo = T, fig.width = 35, fig.height = 35}
bar.grid  <- grid.arrange( as.grob(ctrJ.1),
    as.grob(ctrJ.2),
    as.grob(ctrJ.3),
    as.grob(ba001.BR1),
    as.grob(ba002.BR2),
    as.grob(ba003.BR3),
    ncol = 3,nrow = 2,
    top = textGrob("Barplots for variables", gp = gpar(fontsize = 15, font = 5))
  )

bar.grid 

bar.grid <- recordPlot()

```

```{r}
# Here we can save all figures to a PowerPoint
savedList <- saveGraph2pptx(file2Save.pptx = 'PCA', 
                            title = 'All Figures', 
                            addGraphNames = TRUE)

```

<p style="text-align:center;">## Summary
When we interpret the factor scores and loadings together, the PCA revealed:

* Component 1: Participanta who were grouped as high Autobiographical Memory have higher episodic memory.

* Component 2: Tends to seperate semantic memory which from memory that requires imagination like future and spatial memory.

Note: Semantic memory is based on facts, meanings, concepts and knowledge about the external world that we have acquired and is  independent of personal experience and of the spatial/temporal context in which it was acquired.

* Component 3: Future memory V/s Spatial Memory.</p>

