---
title: "DiSTATIS"
author: "Kirthana S"
date: "06/11/2019"
output: html_document
---

# DiSTATIS

**Data Table:** DiSTATIS method is used when we have K distance matrices collected on the same set of observations.

**Goal:** 

* DISTATIS first evaluates the similarity between distance matrices. 

* From this analysis, a compromise matrix is computed which represents the best aggregate of the original matrices. The
  original distance matrices are then projected onto the compromise matrix.

**Interpretations**
        
    1. Global Factor Scores
       
       The coordinates of the observations on the components are called factor scores and these can be used to plot maps of
       the observations in which the observations are represented as points such that the distances in the map best reflect
       the similarities between the observations.
       
    2. Partial Factor Scores
       
       The positions of the observations ‘as seen by’ each data set are called partial factor scores and can be also
       represented as points in the compromise map. The average of the factor scores of all the tables gives back the
       factor score of the compromise.  



**Note:** Means of the partial factor scores helps to understand the interpretaions better.

## Dataset : Music Composers Dataset 

**Rows:** [composer].[pianist].[ID]
Composers : Beethoven, Bach, Mozart
Pianist : Richter, Arrau, Pires, Baren
Design : color the music or compute means factor scores according to the composers or the conductors

**Columns:** 37 participants 
Design: based on music experience



```{r include=FALSE}

rm(list = ls())
graphics.off()
#devtools::install_github('HerveAbdi/DistatisR',force=TRUE)
# devtools::install_github('HerveAbdi/R4SPISE2018')
#install.packages("DistatisR")
#install.packages("R4SPISE2019")
#install.packages("wesanderson")

#install.packages("dplyr")
#library(R4SPISE2019)

library(dplyr)
library(tidyverse)
library(ExPosition)
library(DistatisR)

library(PTCA4CATA)
library(data4PCCAR)

#install.packages("readxl")
#install.packages("cowplot")
library(ggcorrplot)
library(corrplot)
library(ggplot2)
library(cowplot)
# install.packages('gplots')
library(readr)
library(gridExtra)

library(ggplotify)
library(grid)


# for heatmap2
#install.packages("gplots")
#library(gplots)

knitr::opts_chunk$set(echo = TRUE)
```



```{r include=FALSE}
setwd('C:\\Users\\KIRTHANA\\Desktop\\UT dallas\\Sem3\\RM3\\DiSTATIS')

Raw_Data <- read.csv('natural36_constrained_forR.csv', row.names = 1)
Sorting_Data <- Raw_Data[-37,c(3:39)]
Design_column <- Raw_Data[37, c(3:39)]
Design_row <- Raw_Data[-37,c(1:2)]

#creating the design variable for column
Design_column <- ifelse(Design_column <2,"Novice",     
                  ifelse(Design_column >1 & Design_column <8, "Medium","Expert"))


```

Setting up colours. 

```{r, echo=TRUE}

#color for row design variables
col4Music<- Design_row$Music
col4Music<-dplyr::recode(col4Music, 
                  Bach = "skyblue", 
                  Beet = "olivedrab1",
                  Mozart ="maroon2") 


col4pianist<- Design_row$Pianist
col4pianist<-dplyr::recode(col4pianist,
                    Arrau = "violet",
                    Richt = "violetred",
                    Baren = "slateblue1",
                    Pires = "midnightblue")



#color for column design variables
novice<- Sorting_Data[c(1:17)]
medium<- Sorting_Data[c(18:30)]
expert<- Sorting_Data[c(31:37)]

low <- 'gold'
med <- 'tomato'
high <- 'darkred'
col1 <- rep(low,length(novice))
col2 <- rep(med,length(medium))
col3 <- rep(high, length(expert))

col4col <- as.matrix(c(col1,col2,col3))



#color for group means
col4means<- gplots::col2hex(c( 'darkred','tomato','orange' ))

col4Music.Means<-gplots::col2hex(c( "skyblue", "olivedrab1","maroon2"))

col4pianist.Means<-as.matrix( gplots::col2hex(c( "violet","violetred","slateblue1", "midnightblue")))

```

## DiSTATIS Analysis

Distance cube with each participant representing a matrix which forms a cube when stacked together.
```{r, echo=TRUE}

# Create distance matrices
DistanceCube <- DistatisR::DistanceFromSort(Sorting_Data)

# Run the Plain DiSTATIS analysis
resDistatis <- DistatisR::distatis(DistanceCube)
```

Group Means in the Rv space 
```{r echo=TRUE}

# Get the factors from the Cmat analysis
G <- resDistatis$res4Cmat$G
participant.mean_temp<- aggregate(G, by = list(t(Design_column)), mean)
participant.mean <- participant.mean_temp[,2:ncol(participant.mean_temp )]
rownames(participant.mean)<- participant.mean_temp[,1]


# Get the bootstrap estimates
BootCube <- PTCA4CATA::Boot4Mean(G, design = t(Design_column),
                                 niter = 100,
                                 suppressProgressBar = TRUE)

```

**Looking at the data pattern**

Heat Map of Rv Matrix

RV is a “cosine” between “vectorized” matrices(like correlation)
squared cosine value ranges between: 0 ≤RV ≤1 


```{r}
rvmatrix <- resDistatis$res4Cmat$C
heatmap(rvmatrix, Rowv=NA,Colv = NA)

```

Scree Plots

Rv scree plot - Depicts the homogenity amongst participants. 
```{r, echo=TRUE}
#Scree Plot for Rv matrix
scree.rv <- PlotScree(ev = resDistatis$res4Cmat$eigValues,plotKaiser = TRUE,
                      title = "RV-map: Explained Variance per Dimension")
scree.rv
#Scree plot for the Compromise


scree.comp <- PlotScree(ev = resDistatis$res4Splus$eigValues,plotKaiser = TRUE,
                         title = "Compromise: Explained Variance per Dimension")
scree.comp

```

Column Factor Map :(Participants)

The groups formed are based on the level of experience of the participants.

The novice group, medium and expert group means tend to overlap without a significant separation.
```{r echo=FALSE}

Fj <- resDistatis$res4Cmat$G
colnames(Fj)<- c(paste0('Dimension ',1:37))

Fj.plot<-createFactorMap(Fj,
                         axis1 = 1, axis2 = 2,
                         title = "Participants: RVMap",
                         col.points = col4col,
                         col.labels = col4col)

# create the labels for the dimensions of the RV map
labels4RV <- createxyLabels.gen(lambda = resDistatis$res4Cmat$eigValues ,
                                tau = resDistatis$res4Cmat$tau,
                                axisName = "Dimension ")


# Create the map from the layers
# Here with lables and dots
Fj.1map<- Fj.plot$zeMap + labels4RV
# Here with colored dots only
Fj.2map <- Fj.plot$zeMap_background + 
                Fj.plot$zeMap_dots + 
                labels4RV

Fj.2map

#------------------------------------------------------------------------------------------
#plotting the means of the participant

Design_column<-t(Design_column)

grp.ind <- order(Design_column)[!duplicated(sort(Design_column))]
grp.col <- col4col
grp.name <- Design_column[grp.ind] # get the corresponding groups
names(grp.col) <- grp.name

#------------------------------------------------------------------------------------
group.mean <- aggregate(Fj,
                        by = list(Design_column),
                        # must be a list
                        mean
                        )
rownames(group.mean) <- group.mean[,1] # Use the first column as row names
fj.mean <- group.mean[,-1] # Exclude the first column
fj.mean
#------------------------------------------------------------------------------------------
fj.mean.plot <- createFactorMap(fj.mean,
                                alpha.points = 1,
                                display.labels = TRUE,
                                col.points = col4means,
                                col.labels = col4means,
                                pch = 17,
                                cex = 3,
                                text.cex = 4
                                )

Fj.withMeans <- Fj.2map + labels4RV +fj.mean.plot$zeMap_dots + fj.mean.plot$zeMap_text


Fj.withMeans 

#creating confidence intervals

BootCube <- PTCA4CATA::Boot4Mean(Fj, 
                                 design = Design_column,
                                 niter = 100,
                                 suppressProgressBar = TRUE)



# Create the ellipses using function MakeCIEllipses from package PTCA4CATA
GraphElli <- PTCA4CATA::MakeCIEllipses(BootCube$BootCube[,1:2,],
                                       names.of.factors = c("Dimension 1","Dimension 2"),
                                       col = col4means, 
                                       p.level = .95
)

# create the I-map with Observations, means and confidence intervals
Map.J.withCI <-  Fj.withMeans +  GraphElli + labels4RV + ggtitle(" PArticipant Groups with Confidence Intervals")

Map.J.withCI

```
Partial Factor scores 

Compare the partial factor scores between Novice, Medium, Expert
```{r echo=TRUE}
F_j <- resDistatis$res4Splus$PartialF 
alpha_j <- resDistatis$res4Cmat$alpha

Group_Participant<- (Design_column)
code4Groups <- unique(Group_Participant)
nK <- length(code4Groups)

# initialize F_K and alpha_k
F_k <- array(0, dim = c(dim(F_j)[[1]], dim(F_j)[[2]],nK)) 
dimnames(F_k) <- list(dimnames(F_j)[[1]],
dimnames(F_j)[[2]], code4Groups)
alpha_k <- rep(0, nK)
names(alpha_k) <- code4Groups
Fa_j <- F_j
# A horrible loop
for (j in 1:dim(F_j)[[3]]){ Fa_j[,,j] <- F_j[,,j] * alpha_j[j] }
# Another horrible loop
for (k in 1:nK){
lindex <- Group_Participant == code4Groups[k]
alpha_k[k] <- sum(alpha_j[lindex])
F_k[,,k] <- (1/alpha_k[k])*apply(Fa_j[,,lindex],c(1,2),sum)
}

pFi <- F_k

```





Row Factor Map based on design : Music Pieces


The group means of Bach, Mozart and Beethoven music groups are slighly separated. However their confidence intervals tend to overlap. 
```{r row factor maps, echo=FALSE}


Fi <- resDistatis$res4Splus$F
#constraints4Fi <- lapply(minmaxHelper(Fi),'*',1.2)
#colnames(Fj)<- c(paste0('Dimension ',1:37))

#labels for the compromise plot
# to get the eigen values and tau for the compromise

#color4Products <- # Create color for the Products from prettyGraph
#prettyGraphsColorSelection(n.colors = nrow(resDistatis$res4Splus$F))

Fi.plot <- PTCA4CATA::createFactorMap(Fi,
                                        constraints = minmaxHelper4Partial(Fi, pFi,  
                                        axis1 = 1, axis2 = 2),
                                        col.points = col4Music,
                                        col.labels = col4Music,
                                        axis1 = 1,
                                        axis2 = 2,
                                        title = 'Music Pieces Map',
                                        alpha.points = 0.8,
                                      display.labels = FALSE
)

labels4S <- createxyLabels.gen(lambda = resDistatis$res4Splus$eigValues,
                                tau = resDistatis$res4Splus$tau,
                                axisName = "Dimension ")

# Plain map with color for the I-set
Fi.1map <- Fi.plot$zeMap + labels4S 
  

#Creating means 

design.music<- as.matrix(Design_row$Music)

grp.ind <- order(design.music)[!duplicated(sort(design.music))]
grp.col <- col4Music
grp.name <- design.music[grp.ind] # get the corresponding groups
names(grp.col) <- grp.name

#------------------------------------------------------------------------------------
Music.mean <- aggregate(Fi,
                        by = list(design.music),
                        # must be a list
                        mean
                        )



rownames(Music.mean) <- Music.mean[,1] # Use the first column as row names
fi.mean <- Music.mean[,-1] # Exclude the first column
#fi.mean
#------------------------------------------------------------------------------------------
fi.mean.plot <- createFactorMap(fi.mean,
                                alpha.points = 1,
                                display.labels = TRUE,
                                col.points = col4Music.Means,
                                col.labels = col4Music.Means,
                                pch = 17,
                                cex = 3,
                                text.cex = 4
                                )

Fi.withMeans <- Fi.1map + labels4S +fi.mean.plot$zeMap_dots + fi.mean.plot$zeMap_text


Fi.withMeans 

#creating confidence intervals

BootCube <- PTCA4CATA::Boot4Mean(Fi, 
                                 design = design.music,
                                 niter = 100,
                                 suppressProgressBar = TRUE)



# Create the ellipses using function MakeCIEllipses from package PTCA4CATA
GraphElli <- PTCA4CATA::MakeCIEllipses(BootCube$BootCube[,1:2,],
                                       names.of.factors = c("Factor 1","Factor 2"),
                                       col = col4Music.Means, 
                                       p.level = .95
)

# create the I-map with Observations, means and confidence intervals
Map.J.withCI <-  Fi.withMeans +  GraphElli + labels4S + ggtitle(" Music Pieces with Confidence Intervals")

Map.J.withCI

```

Row Factor Map based on degsign Pianists


The group means of Baren is significantly different from Arrau Richt and Pires whose group means tend to overlap each other.

```{r echo=FALSE}
Fi <- resDistatis$res4Splus$F
#constraints4Fi <- lapply(minmaxHelper(Fi),'*',1.2)
#colnames(Fj)<- c(paste0('Dimension ',1:37))

#labels for the compromise plot
# to get the eigen values and tau for the compromise

#color4Products <- # Create color for the Products from prettyGraph
#prettyGraphsColorSelection(n.colors = nrow(resDistatis$res4Splus$F))

Fi2.plot <- PTCA4CATA::createFactorMap(Fi,
                                       
                                        col.points = col4pianist,
                                        col.labels = col4pianist,
                                        axis1 = 1,
                                        axis2 = 2,
                                        title = 'Music Pieces Map',
                                        alpha.points = 0.8,
                                      display.labels = FALSE
)

labels4S <- createxyLabels.gen(lambda = resDistatis$res4Splus$eigValues,
                                tau = resDistatis$res4Splus$tau,
                                axisName = "Dimension ")

# Plain map with color for the I-set
Fi.2map <- Fi2.plot$zeMap + labels4S 
  



#Creating means 

design.pianist<- as.matrix(Design_row$Pianist)

grp.ind <- order(design.pianist)[!duplicated(sort(design.pianist))]
grp.col <- col4pianist
grp.name <- design.pianist[grp.ind] # get the corresponding groups
names(grp.col) <- grp.name

#------------------------------------------------------------------------------------
pianist.mean <- aggregate(Fi,
                        by = list(design.pianist),
                        # must be a list
                        mean
                        )



rownames(pianist.mean) <- pianist.mean[,1] # Use the first column as row names
fi2.mean <- pianist.mean[,-1] # Exclude the first column

#------------------------------------------------------------------------------------------
fi2.mean.plot <- createFactorMap(fi2.mean,
                                alpha.points = 1,
                                display.labels = TRUE,
                                col.points = col4pianist.Means,
                                col.labels = col4pianist.Means,
                                pch = 17,
                                cex = 3,
                                text.cex = 4
                                )

Fi2.withMeans <- Fi.2map + labels4S +fi2.mean.plot$zeMap_dots + fi2.mean.plot$zeMap_text


Fi2.withMeans 

#creating confidence intervals

BootCube <- PTCA4CATA::Boot4Mean(Fi, 
                                 design = design.pianist,
                                 niter = 100,
                                 suppressProgressBar = TRUE)



# Create the ellipses using function MakeCIEllipses from package PTCA4CATA
GraphElli <- PTCA4CATA::MakeCIEllipses(BootCube$BootCube[,1:2,],
                                       names.of.factors = c("Factor 1","Factor 2"),
                                       col = col4pianist.Means, 
                                       p.level = .95
)

# create the I-map with Observations, means and confidence intervals
Map.J2.withCI <-  Fi2.withMeans +  GraphElli + labels4S + ggtitle(" Music Pieces with Confidence Intervals")

Map.J2.withCI

```

Plotting the partial Factor scores

The experts amongst the participants appear to be much diffferent than the rest of the participants because their dintance from the compromise factor score is larger.

This is much more clear while observing the means of the partial factor scores.

```{r}
pFi <- F_k


map4PFS <- createPartialFactorScoresMap(
                                        factorScores = Fi,
                                        partialFactorScores = pFi,
                                        
                                        axis1 = 1, axis2 = 2,
                                        colors4Items = as.vector(col4Music),
                                        colors4Blocks = as.vector(col4means),
#colors4Blocks = c("lightblue", "skyblue","midnightblue"),

                                        names4Partial = dimnames(pFi)[[3]], #
                                        font.labels = 'bold',
                                        size.labels = 2)

plot.pFi1 <-  Fi.plot$zeMap +map4PFS$mapColByItems 
plot.pFi1 



plot.pFi2 <- Fi.plot$zeMap + labels4S + map4PFS$mapColByBlocks 

plot.pFi2
```

Partial factor scores with Means(Music pieces)


Experts are different from movice by a large amount when trying to group Mozart and Bach music pieces together than when grouping Beethoven pieces together. 

While goruping Beethiven music together, experts and medium level participants do almost an equal job. 

```{r echo=TRUE}
 
meanfk <-
  apply(F_k, c(2,3), FUN = function(x){
  aggregate(x, by = list(Design_row$Music), mean)$x
  })
dim(meanfk)  

mean.plot <- createFactorMap(fi.mean,
                                constraints = minmaxHelper4Partial(fi.mean, meanfk, axis1 = 1 ,axis2 = 2) ,                               
                              alpha.points = 1,
                                display.labels = TRUE,
                                col.points = col4Music.Means,
                                col.labels = col4Music.Means,
                                pch = 17,
                                cex = 3,
                                text.cex = 4
                                )

Fi.meanonly.plot<- mean.plot$zeMap_background+mean.plot$zeMap_dots + mean.plot$zeMap_text+ labels4S

Fi.meanonly.plot


pf.means <- createPartialFactorScoresMap(
                                        factorScores = fi.mean,
                                        partialFactorScores = meanfk,
                                        axis1 = 1, axis2 = 2,
                                        colors4Items = as.vector(col4Music.Means),
colors4Blocks = as.vector(col4means),
                                        names4Partial = dimnames(meanfk)[[3]], #
                                        font.labels = 'bold',
                                        size.labels = 4,
)

plot.pFi.mean <- Fi.meanonly.plot + labels4S+ pf.means$mapColByItems

plot.pFi.mean2 <- Fi.meanonly.plot + labels4S+ pf.means$mapColByBlocks
plot.pFi.mean2

```


Partial factor scores with Means(Pianist)

Experts are different from movice by a large amount when trying to group Arrau and Pires pieces together.
Medium Group is not significantly different in grouping any of these pianinsts.

```{r echo=FALSE}


meanfk2 <-
  apply(F_k, c(2,3), FUN = function(x){
  aggregate(x, by = list(Design_row$Pianist), mean)$x
  })
dim(meanfk2)  


mean.plot2 <- createFactorMap(fi2.mean,
                                constraints = minmaxHelper4Partial(fi2.mean, meanfk2, axis1 = 1 ,axis2 = 2) ,                               
                              alpha.points = 1,
                                display.labels = TRUE,
                                col.points = col4pianist.Means,
                                col.labels = col4pianist.Means,
                                pch = 17,
                                cex = 3,
                                text.cex = 4
                                )

Fi.meanonly.plot2<- mean.plot2$zeMap_background+ mean.plot2$zeMap_dots + mean.plot2$zeMap_text+ labels4S

Fi.meanonly.plot2


pf.means2 <- createPartialFactorScoresMap(
                                        factorScores = fi2.mean,
                                        partialFactorScores = meanfk2,
                                        axis1 = 1, axis2 = 2,
                                        colors4Items = as.vector(col4pianist.Means),
colors4Blocks = as.vector(col4means),
                                        names4Partial = dimnames(meanfk2)[[3]], #
                                        font.labels = 'bold',
                                        size.labels = 4,
)

mean.pfi <- Fi.meanonly.plot2 + labels4S+ pf.means2$mapColByItems


mean.pfi.block <- Fi.meanonly.plot2 + labels4S+ pf.means2$mapColByBlocks
mean.pfi.block



```